Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-08 paper/main.tex
typo

\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\itemskipIn
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver, and does not require any
  fixpoint computations to obtain loop invariants.
\itemskipIn
\item We show with a publicly available prototype implementation, and
  an experimental evaluation, that our novel amortized resource
  analysis works precisely and efficiently for challenging loop and
  recursion patterns, and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
\itemskip
%
To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Furthermore, the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during execution.

@@

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract interpretation--based methods such as disjunctive invariant
generation~\cite{GulwaniMC09} and symbolic backward
execution~\cite{GulwaniZ10}.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects the resource cost and size changes of
integer variables in the program; a solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

@@

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations of variables.  It is used, for example, to determine
at variable assignments if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

@@

\pref{fig:compar} shows five representative loop patterns from the
evaluation.  Example \emph{t08} is a slightly modified version of
\emph{t08a} which is described in the introduction.  Example
\emph{t19} demonstrates the compositionaliy of the analysis.  The
program consists of two loops that decrement a variable $i$.  In the
first loop, $i$ is decremented down to 100 and in the second loop $i$
is decremented further down to $-1$.  However, between the loops we
assign \code{i=i+k+50}.  So in total the program performs $52 + |[-1,i]| +
|[0,k]|$ increments.  Our analysis finds this tight bound because our
amortized analysis naturally takes into account the relation between
the two loops.

>>>>

\item We develop a quantitative Hoare logic for interactively deriving
  resource bounds for Clight programs.  The logic is implemented and
  proved sound in the Coq Proof Assistant with respect to the cost
  semantics of Clight.
\itemskipIn
\item We describe an automatic static analysis that computes bounds
  together with derivations in the quantitative Hoare logic.  The
  analysis is inspired by type-based amortized resource analysis for
  functional programs~\cite{Jost03,HoffmannAH11}.  It computes
  derivations by generating simple linear constraints that can be
  solved by an off-the-shelf LP solver, and does not require any
  fixpoint computations to obtain loop invariants.
\itemskipIn
\item We show with a publicly available prototype implementation, and
  an experimental evaluation, that our novel amortized resource
  analysis works precisely and efficiently for challenging loop and
  recursion patterns, and that the derived constant factors in the
  bounds are close to or identical with the optimal ones.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{enumerate}
\itemskip
%
To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still supporting
automation that sometimes goes beyond the capabilities of existing
techniques.  Furthermore, the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during execution.

@@

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract interpretation--based methods such as disjunctive invariant
generation~\cite{GulwaniMC09} and symbolic backward
execution~\cite{GulwaniZ10}.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects the resource cost and size changes of
integer variables in the program; a solution of the linear program
immediately yields a resource-usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

@@

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { tick(-1); x=x+1; tick(1); } #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
  while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
More formally, we develop (in \pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations of variables.  It is used, for example, to determine
at variable assignments if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

@@

\pref{fig:compar} shows five representative loop patterns from the
evaluation.  Example \emph{t08} is a slightly modified version of
\emph{t08a}, which is described in~\pref{sec:inform}.  Example
\emph{t19} demonstrates the compositionaliy of the analysis.  The
program consists of two loops that decrement a variable $i$.  In the
first loop, $i$ is decremented down to 100 and in the second loop $i$
is decremented further down to $-1$.  However, between the loops we
assign \code{i=i+k+50}.  So in total the program performs $52 + |[-1,i]| +
|[0,k]|$ increments.  Our analysis finds this tight bound because our
amortized analysis naturally takes into account the relation between
the two loops.
